---
title: "Homework 4"
author: "Jordan Nieusma"
date: "2023-02-27"
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library('MASS') ## for 'mcycle'
library('manipulate') ## for 'manipulate'
library(tidyr)
```

- [X] Randomly split the mcycle data into training (75%) and validation (25%) subsets.


```{r}
set.seed(1)

sample <- sample(c(TRUE,FALSE), nrow(mcycle), 
                 replace=TRUE, prob=c(0.75,0.25))
train <- mcycle[sample, ]
test <- mcycle[!sample, ]
```

- [X] Using the mcycle data, consider predicting the mean acceleration as a function of time. 

```{r}
## set y = accel and X = times
X_train <- train$times
X_test <- test$times
y_train <- train$accel
y_test <- test$accel

fit <- lm(accel ~ times, data=train)
fit
```

- [] Use the Nadaraya-Watson method with the k-NN kernel function to create a series of prediction models by varying the tuning parameter over a sequence of values. (hint: the script already implements this)

```{r}
## Make predictions using the NW method
## y  - n x 1 vector of training outputs
## x  - n x p matrix of training inputs
## x0 - m x p matrix where to make predictions
## kern  - kernel function to use
## ... - arguments to pass to kernel function
nadaraya_watson <- function(y, x, x0, kern, ...) {
  k <- t(apply(x0, 1, function(x0_) {
    k_ <- kern(x, x0_, ...)
    k_/sum(k_)
  }))
  yhat <- drop(k %*% y)
  attr(yhat, 'k') <- k
  return(yhat)
}

## Helper function to view kernel (smoother) matrix
matrix_image <- function(x) {
  rot <- function(x) t(apply(x, 2, rev))
  cls <- rev(gray.colors(20, end=1))
  image(rot(x), col=cls, axes=FALSE)
  xlb <- pretty(1:ncol(x))
  xat <- (xlb-0.5)/ncol(x)
  ylb <- pretty(1:nrow(x))
  yat <- (ylb-0.5)/nrow(x)
  axis(3, at=xat, labels=xlb)
  axis(2, at=yat, labels=ylb)
  mtext('Rows', 2, 3)
  mtext('Columns', 3, 3)
}

## Compute effective df using NW method
## y  - n x 1 vector of training outputs
## x  - n x p matrix of training inputs
## kern  - kernel function to use
## ... - arguments to pass to kernel function
effective_df <- function(y, x, kern, ...) {
  y_hat <- nadaraya_watson(y, x, x,
    kern=kern, ...)
  sum(diag(attr(y_hat, 'k')))
}

## loss function
## y    - train/test y
## yhat - predictions at train/test x
loss_squared_error <- function(y, yhat)
  (y - yhat)^2

## test/train error
## y    - train/test y
## yhat - predictions at train/test x
## loss - loss function
error <- function(y, yhat, loss=loss_squared_error)
  mean(loss(y, yhat))

## AIC
## y    - training y
## yhat - predictions at training x
## d    - effective degrees of freedom
aic <- function(y, yhat, d)
  error(y, yhat) + 2/length(y)*d

## BIC
## y    - training y
## yhat - predictions at training x
## d    - effective degrees of freedom
bic <- function(y, yhat, d)
  error(y, yhat) + log(length(y))/length(y)*d


## make predictions using NW method at training inputs
y_hat <- nadaraya_watson(y_train, X_train, X_train,
  kernel_epanechnikov, lambda=5)

## view kernel (smoother) matrix
matrix_image(attr(y_hat, 'k'))

## compute effective degrees of freedom
edf <- effective_df(y, x, kernel_epanechnikov, lambda=5)
aic(y, y_hat, edf)
bic(y, y_hat, edf)

## create a grid of inputs 
x_plot <- matrix(seq(min(x),max(x),length.out=100),100,1)

## make predictions using NW method at each of grid points
y_hat_plot <- nadaraya_watson(y, x, x_plot,
  kernel_epanechnikov, lambda=1)

## plot predictions
plot(x, y, xlab="Time (ms)", ylab="Acceleration (g)")
lines(x_plot, y_hat_plot, col="#882255", lwd=2) 

## how does k affect shape of predictor and eff. df using k-nn kernel ?
# manipulate({
#   ## make predictions using NW method at training inputs
#   y_hat <- nadaraya_watson(y, x, x,
#     kern=kernel_k_nearest_neighbors, k=k_slider)
#   edf <- effective_df(y, x, 
#     kern=kernel_k_nearest_neighbors, k=k_slider)
#   aic_ <- aic(y, y_hat, edf)
#   bic_ <- bic(y, y_hat, edf)
#   y_hat_plot <- nadaraya_watson(y, x, x_plot,
#     kern=kernel_k_nearest_neighbors, k=k_slider)
#   plot(x, y, xlab="Time (ms)", ylab="Acceleration (g)")
#   legend('topright', legend = c(
#     paste0('eff. df = ', round(edf,1)),
#     paste0('aic = ', round(aic_, 1)),
#     paste0('bic = ', round(bic_, 1))),
#     bty='n')
#   lines(x_plot, y_hat_plot, col="#882255", lwd=2) 
# }, k_slider=slider(1, 10, initial=3, step=1))
```


- [] With the squared-error loss function, compute and plot the training error, AIC, BIC, and validation error (using the validation data) as functions of the tuning parameter.

- [] For each value of the tuning parameter, Perform 5-fold cross-validation using the combined training and validation data. This results in 5 estimates of test error per tuning parameter value.

- [] Plot the CV-estimated test error (average of the five estimates from each fold) as a function of the tuning parameter. Add vertical line segments to the figure (using the segments function in R) that represent one “standard error” of the CV-estimated test error (standard deviation of the five estimates from each fold).

- [] Interpret the resulting figures and select a suitable value for the tuning parameter.
