---
title: "Homework 2"
author: "Jordan Nieusma"
date: "January 19, 2023"
output: github_document
---

## Setup:

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
library(qrnn)
library(dplyr)
```

```{r load-data}
dat <- 
  read.table(url(
    'https://web.stanford.edu/~hastie/ElemStatLearn/datasets/prostate.data'))
# subset to training examples
dat_train <- subset(dat, train==TRUE)
```

## Write code that implements L2, L1, and tilted absolute loss functions

```{r loss-fns}
# L2 loss function
L2_loss <- function(y, yhat)
  (y-yhat)^2

# L1
L1_loss <- function(y, yhat) {
  mean(abs(y - yhat))
}

# Tilted loss, tau=0.25
tilt75_loss <- function(y, yhat, tau) {
  qrnn::tilted.abs(y-yhat, tau=0.25)
}

# Tilted loss, tau=0.75
tilt25_loss <- function(y, yhat)
  qrnn::tilted.abs(y-yhat, tau = 0.75)

```

## Plot lpsa (x-axis) versus lcavol (y-axis). 

```{r plot-actuals}
# Create a data frame for plotting actual values
df <- data.frame(x=dat$lpsa, y=dat$lcavol)

# Plot actuals (lpsa and lcavol)
plot_psa_data <- function(dat=dat_train) {
  plot(dat$lpsa, dat$lcavol,
       xlab="log Prostate Screening Antigen (psa)",
       ylab="log Cancer Volume (lcavol)",
       pch = 20)
}
plot_psa_data()
```

## Add and label the linear model predictors associated with L2 loss, L1 loss, and tilted absolute value loss for tau = 0.25 and 0.75.

```{r fit-pred-fns}
# Write functions to fit and predict on the training data
# Fit
fit_lin <- function(y, x, loss, beta_init = c(-0.51, 0.75)) {
  err <- function(beta)
    mean(loss(y, beta[1] + beta[2]*x))
  beta <- optim(par=beta_init, fn=err)
  return(beta)
}

# Predict
predict_lin <- function(x, beta)
  beta[1] + beta[2]*x

# Compute predictions for a grid of inputs
x_grid <- seq(min(dat_train$lpsa),
              max(dat_train$lpsa),
              length.out=100)
```


```{r plot-lin-preds}
# Fit and predict using each loss fn
# L1
lin_beta_L1 <- fit_lin(y=dat_train$lcavol, x=dat_train$lpsa, loss=L1_loss)
lin_pred_L1 <- predict_lin(x=x_grid, beta=lin_beta_L1$par)

# L2
lin_beta_L2 <- fit_lin(y=dat_train$lcavol, x=dat_train$lpsa, loss=L2_loss)
lin_pred_L2 <- predict_lin(x=x_grid, beta=lin_beta_L2$par)

# Abs tilted loss
# tau = 0.25
lin_beta_tilt25 <- fit_lin(y=dat_train$lcavol, x=dat_train$lpsa, loss=tilt25_loss)
lin_pred_tilt25 <- predict_lin(x=x_grid, beta=lin_beta_tilt25$par)

# tau = 0.75
lin_beta_tilt75 <- fit_lin(y=dat_train$lcavol, x=dat_train$lpsa, loss=tilt75_loss)
lin_pred_tilt75 <- predict_lin(x=x_grid, beta=lin_beta_tilt75$par)


# Add predictions to plot
plot_psa_data()
lines(x=x_grid, y=lin_pred_L1, col='blue', lwd=2, lty=1)
lines(x=x_grid, y=lin_pred_L2, col='pink', lwd=2, lty=1)
lines(x=x_grid, y=lin_pred_tilt25, col='purple', lwd=2, lty=1)
lines(x=x_grid, y=lin_pred_tilt75, col='red', lwd=2, lty=1)

legend(-0.5, 4, legend=c("L1 Loss", "L2 Loss", "Tilted Abs Loss (tau=0.25)", "Tilted Abs Loss (tau=0.75)"),
                         col=c("darkgreen", "pink", "purple", "red"),lty=c(1,1,1,1), cex=0.8)
```


## Write functions to fit and predict from a simple nonlinear model with three parameters defined by 'beta[1] + beta[2]*exp(-beta[3]*x)'.

```{r nonlin-fns}
# Non-linear model
fit_nonlin <- function(y, x, loss=L2_loss, beta_init = c(-1.0, 0.0, -0.3 )) {
  err <- function(beta)
    mean(loss(y,  beta[1] + beta[2]*exp(-beta[3]*x)))
  beta <- optim(par = beta_init, fn = err)
  return(beta)
}

predict_nonlin <- function(x, beta)
  beta[1] + beta[2]*exp(-beta[3]*x)

nonlin_beta <- fit_nonlin(y=dat_train$lcavol,
                    x=dat_train$lpsa,
                    loss=L2_loss)

preds_nonlin <- predict_nonlin(x=x_grid, beta=nonlin_beta$par)

```

## Create a figure that shows lpsa (x-axis) versus lcavol (y-axis). Add and label the nonlinear model predictors associated with L2 loss, L1 loss, and tilted absolute value loss for tau = 0.25 and 0.75.


```{r nonlin-preds}
# L2 loss 
nlin_beta_L2 <- fit_nonlin(y=dat_train$lcavol,
                x=dat_train$lpsa,
                loss=L2_loss)

nlin_pred_L2 <- predict_nonlin(x=x_grid, beta=nlin_beta_L2$par)

# L1 loss 
nlin_beta_L1 <- fit_nonlin(y=dat_train$lcavol,
                x=dat_train$lpsa,
                loss=L1_loss)

nlin_pred_L1 <- predict_nonlin(x=x_grid, beta=nlin_beta_L1$par)

# tilted tau=0.25
nlin_beta_tilt25 <- fit_nonlin(y=dat_train$lcavol,
                    x=dat_train$lpsa,
                    loss=tilt25_loss)

nlin_pred_tilt25 <- predict_nonlin(x=x_grid, beta=nlin_beta_tilt25$par)

# fit linear model with tilted tau=0.75

nlin_beta_tilt75 <- fit_nonlin(y=dat_train$lcavol,
                x=dat_train$lpsa,
                loss=tilt75_loss)

nlin_pred_tilt75 <- predict_nonlin(x=x_grid, beta=nlin_beta_tilt75$par)

```

```{r plot-nonlin}
plot_psa_data()

lines(x=x_grid, y=nlin_pred_L1, col='blue', lwd=1)
lines(x=x_grid, y=nlin_pred_L2, col='pink', lwd=2, lty=1)
lines(x=x_grid, y=nlin_pred_tilt25, col='purple', lwd=2, lty=1)
lines(x=x_grid, y=nlin_pred_tilt75, col='red', lwd=2, lty=1)

legend(-0.5, 3.9, legend=c("L1 Loss", "L2 Loss", "Tilted Abs Loss (tau=0.25)", "Tilted Abs Loss (tau=0.75)"),
                         col=c("blue", "pink", "purple", "red"), lty=c(1,1,1,1), cex=0.8)
```


