---
title: "Homework 5"
author: "Jordan Nieusma"
date: "2023-03-30"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Set up the notebook 

- Import packages (dplyr, randomForest, caret)

- Read in test and training data from the vowel dataset at https://hastie.su.domains/ElemStatLearn/datasets

```{r set-up, include=FALSE}
# Packages
library(dplyr)
library(randomForest)
library(caret)

# Load datasets
url_train <- "https://hastie.su.domains/ElemStatLearn/datasets/vowel.train"
url_test <- "https://hastie.su.domains/ElemStatLearn/datasets/vowel.test"
train <- read.csv(url_train, row.names = 1)
test <- read.csv(url_test, row.names = 1)
```

## Goal: Understand and implement a random forest classifier.

## Task #1: 

Using the “vowel.train” data, develop a random forest (e.g., using the "randomForest" package) for the vowel data, using all of the 11 features and using the default values of the tuning parameters.


```{r t1}
# Prepare train/test dfs for tasks
train$y <- factor(train$y)
test$y <- factor(test$y)

rf_default <- randomForest(y ~ ., data=train)
print(rf_default)
```

## Task #2: 

Use 5-fold CV to tune the number of variables randomly sampled as candidates at each split if using random forest, or the ensemble size if using gradient boosting.

```{r}
set.seed(1)

# Define seq of values representing no. of variables to try
mtry <- data.frame(mtry=c(1:10))

# Set train method to cross-validation w/5 folds, apply grid search
ctrl <- trainControl(method="cv", number=5, search="grid")

# Train using CV and print results
rf_cv <- train(y ~ ., data=train, method="rf", trControl=ctrl, tuneGrid=mtry)
print(rf_cv)
print(rf_cv$finalModel)
```

## Task #3: 

With the tuned model, make predictions using the majority vote method, and compute the misclassification rate using the ‘vowel.test’ data.

```{r}
# Get X and y variables
X_test <- test[, -c(1)]
y_test <- test[, c(1)]

# Get predictions
pred_prob <- predict(rf_cv, newdata=X_test, type="raw")

# Print results
xtab <- table(pred_prob, y_test)
print(confusionMatrix(xtab))
```
